---
title: "Visualizations in R using the tidyverse"
subtitle: "Sigma Xi Fall Workshop, 2023"
author: "P. Lombardo & J. Kaufman"
output:
  html_document:
    df_print: paged
  html_notebook: default
---

```{r set-up, warning=FALSE, message=FALSE}
library(tidyverse)
library(gapminder)
library(patchwork)
library(RColorBrewer)
```


# Case Study 1: Incrementally building plots.
```{r}
ggplot(data = iris,
       aes(x = Petal.Width,
           y = Sepal.Width))+
    geom_point(aes(color=Species),
               alpha=.7)+
    scale_color_brewer("Species\nof Iris",palette="Set2")+
    scale_x_continuous(limits=c(0,3),
                       breaks=seq(0,3,by=.5))+
    scale_y_continuous(limits=c(0,5))+
    # geom_smooth(method = "lm", se = F)+
    geom_smooth(aes(color=Species), method = "lm", se = F)+
    labs(x="Petal Width (in)",
         y="Sepal Width (in)",
         title="Comparing Sepal and Petal Width among three species of Iris")+
    theme_bw()
```


# Case Study 2: Other plots, and `patchwork`-ing them together
Dropping the diamond data set in favor of spotify?

```{r}
df2<-read.csv('data/spotify-2023.csv',header=T)
# df2 %>% names()

df2<-df2 %>% select(track_name,artist.s._name,released_year,
                    in_apple_charts,
                    danceability_., mode,streams)
```

[Patchwork example?](https://patchwork.data-imaginist.com/articles/guides/layout.html)

**Plot 1: Histogram Example**:
```{r}
p1<-df2 %>%
    ggplot(data = .,
           aes(x=danceability_.))+
    geom_histogram(bins=10,
                   aes(fill=mode),alpha=.6,
                   position=position_dodge(width=3))+
    scale_fill_brewer(palette="Set1",guide="none")+
    labs(color="Key Signature Mode",
         y="Counts",
         x="Danceability Score")+
    theme_bw()
```

**Plot 2: Fancy Violin Plot Example**:
```{r}
p2<-df2 %>%
    arrange(desc(streams)) %>%
    mutate(aboveMedian = ifelse(streams > median(streams),"Yes","No")) %>%
    ggplot(data =.,
           aes(x = aboveMedian,
               y=danceability_.))+
    geom_violin(fill='gray',alpha=.6)+
    geom_point(aes(color=mode),
               alpha=.4,
               position = position_jitter(width=.25))+
    scale_color_brewer(palette="Set1")+
    labs(color="Key Signature Mode",
         y="Danceability Score",
         x="Above Median Stream Counts")+
    theme_bw()+
    theme(legend.position="top")
    
```

**Plot 3: Bar Plot Example**:
```{r}
p3<-df2 %>%
    mutate(aboveMedian = ifelse(streams > median(streams),"Yes","No")) %>%
    ggplot(data =.,
           aes(x=mode,
               fill = mode))+
    geom_bar(aes(y=after_stat(count)/sum(after_stat(count))),
             alpha=.75)+
    scale_fill_brewer("Key Signature\nMode",palette="Set1",guide="none")+
    labs(x="Key Signature Mode",
         y="Percentage")+
    theme_bw()
```

Simple patchwork options:
```{r}
(p1+p2)/p3
```

More advanced layouts:
```{r}
layout <- "
AAAA##
BBBBCC
BBBBCC
"
p1+p2+p3 +plot_layout(design=layout)
```


# Case Study 3: Summarizing and plotting
The gapminder data set looks at metrics like life expectancy and national GDP for several nations around the world over time. 

Let's make a plot consider trends in life expectancy over time by continent.
```{r, warning=FALSE, message=FALSE}
gapminder %>%
    group_by(continent,year) %>%
    summarise(meanPop = mean(pop),
              meanLife = mean(lifeExp),
              sdLife = sd(lifeExp),
              lowLife = meanLife -sdLife,
              highLife = meanLife + sdLife) %>%
    ggplot(data = .,
           aes(x=year,
               y = meanLife,
               color=continent,
               group=continent))+
    geom_point(aes(size = meanPop))+geom_line()+
    # could make the size of the points sdLife.
    scale_color_brewer("Continent",palette="Set2")+
    # geom_errorbar(aes(ymin=lowLife,ymax=highLife),
    #               alpha=.3,width=1)+
    # or use ribbons to show variation
    geom_ribbon(aes(ymin=lowLife,ymax=highLife,fill=continent),
                alpha=.1)+
    scale_fill_brewer("Continent",palette="Set2", guide = "none")+
    scale_x_continuous(breaks=seq(1950,2010,by=5))+
    geom_hline(yintercept=seq(40,80,by=10),
               col='gray',alpha=.3)+
    theme_classic()
```
(*Note:* A bit noisy for my taste....)


# Case Study 4: Non-tidy data

```{r}
# From five-thirty-eight
# https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age
df4<-read.csv('data/drug-use-by-age.csv',header=T,
              na.strings = "-")
# # df4 %>% names()
# df4 %>% str()
```

Columns display same measurements (use as percentage used in 12-month period, frequency as a median number of times in 12-month period) but categories are split among columns.

One tidy format would have `use` and `frequency` as their own columns, with `drug_type` as indicating the type of drug.

```{r}
df4_final<-df4 %>% pivot_longer(
    col = -c(age,n),
    names_pattern = "(.*)_(.*)",
    names_to = c("drug_type","measure_type")
) %>% 
    pivot_wider(
        names_from=measure_type,
        values_from = value
    )

# df4_final$drug_type %>% unique()
# ?pivot_longer
# ?pivot_wider
```

```{r}
display.brewer.all()
```

```{r}
df4_final %>%
    # filter(drug_type %in% c("alcohol","marijuana","meth")) %>%
    ggplot(data =.,
           aes(x = age,
               y=drug_type,
               fill=use))+
    geom_tile()+
    scale_fill_gradientn("Percentage of\nAge Group Using\nin the past year",colors = brewer.pal(9,name="YlOrRd"))+
    theme_dark()+
    theme(axis.text.x = element_text(angle = 60, hjust=1))+
    labs(x="Age Group",
         y="Drug Type",
         title="Drug use percentages by type")
```


