---
title: "Visualizations in R using the tidyverse"
subtitle: "Sigma Xi Fall Workshop, 2023"
author: "P. Lombardo & J. Kaufman"
output:
  html_document:
    df_print: paged
  html_notebook: default
---

Begin by loading the necessary packages and necessary data sets.
```{r set-up, warning=FALSE, message=FALSE}
library(tidyverse)
library(gapminder)
library(patchwork)
library(RColorBrewer)

# Courtesy of Kaggle
spotify<-read.csv('data/spotify-2023.csv',header=T)
    # Select only certain variables
spotify<-spotify %>%
    select(track_name,artist.s._name,
           released_year,
           in_apple_charts,
           danceability_., 
           mode, streams)

# Courtesy of five-thirty-eight
drugUse<-read.csv('data/drug-use-by-age.csv',
                  header=T,
                  na.strings = "-")
```


# Workshop Examples

## Case Study 1: Incrementally building plots with the `iris` data set.
```{r}
head(iris)
```


## Case Study 2: Using `spotify` data to explore plot options and `patchwork` them together

```{r}
head(spotify) %>% as_tibble()
```


**Plot 1: Histogram Example**.
```{r}
# Let's build a histogram together
# p1
```

**Plot 2: Fancy Violin Plot Example**.
```{r}
# What's a violin plot?
# p2
```

**Plot 3: Bar Plot Example**.
```{r}
# how about a simple bar plot...
# p3
```

Simple patchwork options:
```{r}
# (p1+p2)/p3
```

More advanced layouts:
```{r}
layout <- "
AAAA##
BBBBCC
BBBBCC
"
# p1+p2+p3 +plot_layout(design=layout)
```


## Case Study 3: Plotting summary data with the `gapminder` data set

The gapminder data set looks at metrics like life expectancy and national GDP for several nations around the world over time. 

Let's make a plot consider trends in life expectancy over time by continent.
```{r, warning=FALSE, message=FALSE}
head(gapminder)
```


## Case Study 4: Reshaping and plotting with the `drugUse` data set.

```{r}
head(drugUse)
```

Columns display the same measurements (`use` as percentage of the age group that used in the 12-month period, and `frequency` as the median number of times an individual used in the 12-month period), but categories are split among columns.

One tidy format would have `use` and `frequency` as their own columns, with `drug_type` as indicating the type of drug.

```{r}
?pivot_longer
?pivot_wider
```

```{r}
# display.brewer.all()
```

```{r}
# Let's create a heatmap
```

# Helpful Resources:

## Data Repositories

* [Kaggle data sets](https://www.kaggle.com/datasets)
    * [Spotify songs 2023 (Kaggle)](https://www.kaggle.com/datasets/nelgiriyewithana/top-spotify-songs-2023)
* [five-thirty-eight data (github)](https://github.com/fivethirtyeight/data)
    * [druge-use data set](https://github.com/fivethirtyeight/data/tree/master/drug-use-by-age)

## Package documentation

* [`tidyverse` documentation](https://www.tidyverse.org/)
    * [`ggplot` documentation](https://ggplot2.tidyverse.org/)
    * [`dplyr` documentation](https://dplyr.tidyverse.org/)
* [`patchwork` documentation](https://patchwork.data-imaginist.com/)
* [`gapminder` R package documentation](https://cran.r-project.org/web/packages/gapminder/index.html)
    * [Gapminder organization](https://www.gapminder.org/)


